---
title: "EDS 223 Assignment #4"
subtitle: "Prioritizing potential aquaculture on the West Coast of the US"
author: "Maxwell Pepperdine"
date: last-modified
execute: 
  eval: true
  warning: false
  message: false
format:
  html:
    toc: true
    theme: flatly
editor_options: 
  chunk_output_type: console
---

## Load packages

```{r}
#| code-fold: true 
#| code-summary: "Show packages used in this analysis"

library(tidyverse)
library(here)
library(sf)
library(terra)
library(tmap)
```

## Background

With an ever growing human population, marine aquaculture has the potential to play a significant role global food supply, and is a more sustainable option than land-based meat production (Hall et al., 2011). A study that mapped the global potential for marine aquaculture based on multiple constraints (e.g., ship traffic, dissolved oxygen, bottom depth) found that global seafood demand could be met using less than 0.015% of the global ocean area (Gentry et al., 2017). 

This analysis aims to determine which Exclusive Economic Zones (EEZ) on the West Coast of the US are best suited to develop marine aquaculture for several species of oysters and Pacific littleneck clams (*Leukoma staminea*),two of the most common species farmed in marine aquaculture on the West Coast. Suitable locations were determined based on range of suitable sea surface temperature (SST) and depth values for each species listed below. Oyster conditions were provided in the assignment description, and suitable temperatures and depths for the littleneck clam were selected based on Shaw (1986) and Harbo (1997). 

Suitable growing conditions for oysters:

- `sea surface temperature`: 11-30°C
- `depth`: 0-70 meters below sea level

Suitable growing conditions for shrimp:

- `sea surface temperature`: 0-25°C
- `depth`: 0-46 meters below sea level

## Data description

#### Sea surface temperature (SST)

...

#### Bathyemtry

...

#### Exclusive Economic Zones (EEZ)

...

## Analysis

*Pseudo code:*

1. Prepare data. Load all necessary data and make sure they're all in the same CRS. Combine the SST data into a raster stack. If not, transform everything to WGS 84 (`EPSG:4326`).

2. Process data. Find the mean SST from 2008-2012, convert it from Kelvin to Celsius, resample the bathy raster to match the resolution, extent, and position of the SST raster, and crop the depth raster to match the extent of the SST raster.

3. Find suitable locations. Determine which locations meet the suitable growing conditions for oysters and littleneck clams.

#### Load data

```{r}
#| code-fold: true 
#| code-summary: "Show the code"

# shapefile for West Coast EEZ
wc_eez <- st_read(here("data/wc_regions_clean.shp"))

# bathymetry raster
bathy <- rast(here("data/depth.tif"))

# SST rasters
sst_2008 <- rast(here("data/average_annual_sst_2008.tif"))
sst_2009 <- rast(here("data/average_annual_sst_2009.tif"))
sst_2010 <- rast(here("data/average_annual_sst_2010.tif"))
sst_2011 <- rast(here("data/average_annual_sst_2011.tif"))
sst_2012 <- rast(here("data/average_annual_sst_2012.tif"))

# SST raster stack 
sst_stack <- c(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)
#plot(sst_stack)
```

#### Check the CRS of each dataset & transform to that of the `wc_eez` shapefile if they don't match

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# # check if each dataset has the same CRS
# crs(wc_eez) == crs(bathy)
# crs(wc_eez) == crs(sst_stack)
# crs(bathy) == crs(sst_stack)

# none are in the same CRS; transform all to WGS 84
bathy <- project(bathy, 
                 "EPSG:4326")
sst_stack <- project(sst_stack, 
                     "EPSG:4326")
wc_eez <- st_transform(wc_eez, 
                       crs = st_crs(bathy))

# check if they are now in the same CRS
if(crs(wc_eez) == crs(bathy) & crs(wc_eez) == crs(sst_stack)) {
  print("Carry on! All datasets are now in the same CRS") 
} else {
  warning("STOP! Datasets are not in the same CRS")
}
```

#### Process data

```{r}
# find the mean SST from 2008-2012
# create a single raster of average SST
sst_mean <- mean(sst_stack)
# plot(sst_mean) # QC to make sure this created one raster


# convert average SST from Kelvin to Celsius
# subtract 273.15 from each grid cell in the raster
sst_mean_celsius <- sst_mean - 273.15
# plot(sst_mean_celsius) # QC to make sure this worked


# resample the bathy raster 
# match the resolution, extent, and position of the SST raster
bathy_resample <- resample(bathy, 
                           sst_mean_celsius, 
                           method = "near")


# crop the depth raster to match the extent of the SST raster 
bathy_crop <- crop(bathy_resample, 
                   sst_mean_celsius, 
                   mask = TRUE)
```

#### QC: Check that the depth & SST data match in resolution, extent, and CRS

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# resolution 
if(all(res(bathy_crop) == res(sst_mean_celsius))) {
  print("The bathy and SST have the same resolution.")
} else {
  warning("Stop! Resolution does not match.")
}

# extent
if(ext(bathy_crop) == ext(sst_mean_celsius)) {
  print("The bathy and SST raster have the same extent.")
} else {
  warning("STOP! The bathy and SST raster do not have the same extent.")
}

# CRS
if(crs(bathy_crop) == crs(sst_mean_celsius)) {
  print("The bathy and SST data have the same CRS.")
} else {
  warning("STOP! The bathy and SST data do not have the same CRS.")
}
```

#### Plot the bathymetric and SST data

```{r}
#| label: fig-data
#| fig-cap: "Bathymetric and sea surface temperature (SST) data for the West Coast of the US used in the analysus."
#| code-fold: true
#| code-summary: "Show the code"

map1 <- tm_graticules(lines = FALSE) + 
tm_shape(bathy_crop) + 
  tm_raster(palette = "Blues", 
            breaks = c(-4000, -2000, -1000, -500, -100, -50, 0), 
            title = "Bathymetry (m)") + 
tm_shape(wc_eez) + 
  tm_borders(col = "black", 
              lwd = 1.5) + 
  tm_layout(legend.outside = TRUE, 
            frame = FALSE)
map1
```

#### Find suitable locations

```{r}

```

#### Determine the most suitable EEZ

```{r}

```

## Reflection

...

## References

Gentry, R. R., Froehlich, H. E., Grimm, D., Kareiva, P., Parke, M., Rust, M., Gaines, S. D., & Halpern, B. S. Mapping the global potential for marine aquaculture. Nature Ecology & Evolution, 1, 1317-1324 (2017).

Hall, S. J., Delaporte, A., Phillips, M. J., Beveridge, M. & O’Keefe, M. Blue Frontiers: Managing the Environmental Costs of Aquaculture (The WorldFish Center, Penang, Malaysia, 2011).

Harbo, R. M. (1997). Shells & shellfish of the Pacific northwest: a field guide. Madiera Park, BC: Harbour Publishing.

Shaw, W N. (1986) Species profiles: life histories and environmental requirements of coastal fishes and invertebrates (Pacific Southwest): Common littleneck clam. [Protothaca staminea]. United States.















