---
title: "EDS 223 Assignment #4"
subtitle: "Prioritizing potential aquaculture on the West Coast of the US"
author: "Maxwell Pepperdine"
date: last-modified
execute: 
  eval: true
  warning: false
  message: false
format:
  html:
    toc: true
    theme: flatly
editor_options: 
  chunk_output_type: console
---

## Load packages

```{r}

```

## Background

With an ever growing human population, marine aquaculture has the potential to play a significant role global food supply, and is a more sustainable option than land-based meat production (Hall et al., 2011). A study that mapped the global potential for marine aquaculture based on multiple constraints (e.g., ship traffic, dissolved oxygen, bottom depth) found that global seafood demand could be met using less than 0.015% of the global ocean area (Gentry et al., 2017). 

This analysis aims to determine which Exclusive Economic Zones (EEZ) on the West Coast of the US are best suited to develop marine aquaculture for several species of oysters and Pacific littleneck clams (*Leukoma staminea*),two of the most common species farmed in marine aquaculture on the West Coast. Suitable locations were determined based on range of suitable sea surface temperature (SST) and depth values for each species listed below. Oyster conditions were provided in the assignment description, and suitable temperatures and depths for the littleneck clam were selected based on Shaw (1986) and Harbo (1997). 

Suitable growing conditions for oysters:

- `sea surface temperature`: 11-30°C
- `depth`: 0-70 meters below sea level

Suitable growing conditions for shrimp:

- `sea surface temperature`: 0-25°C
- `depth`: 0-46 meters below sea level

## Data description

#### Sea surface temperature (SST)

...

#### Bathyemtry

...

#### Exclusive Economic Zones (EEZ)

...

## Analysis

#### Load data

```{r}

```

#### Prepare data

```{r}

```

#### Process data

```{r}

```

#### Find suitable locations

```{r}

```

#### Determine the most suitable EEZ

```{r}

```

## Reflection

...

## References

Gentry, R. R., Froehlich, H. E., Grimm, D., Kareiva, P., Parke, M., Rust, M., Gaines, S. D., & Halpern, B. S. Mapping the global potential for marine aquaculture. Nature Ecology & Evolution, 1, 1317-1324 (2017).

Hall, S. J., Delaporte, A., Phillips, M. J., Beveridge, M. & O’Keefe, M. Blue Frontiers: Managing the Environmental Costs of Aquaculture (The WorldFish Center, Penang, Malaysia, 2011).

Harbo, R. M. (1997). Shells & shellfish of the Pacific northwest: a field guide. Madiera Park, BC: Harbour Publishing.

Shaw, W N. (1986) Species profiles: life histories and environmental requirements of coastal fishes and invertebrates (Pacific Southwest): Common littleneck clam. [Protothaca staminea]. United States.















